---
title: "**MATH 3190 Homework 6**"
subtitle: |
  Due March 5, 2026\
  Focus: Notes 7 Part 2\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document or pdf document (html is quicker to render and you will need an installation of LaTeX to render as a pdf). You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_6` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this (i.e. don't manually upload files to GitHub). In addition, you will submit just your html or pdf file on Canvas to unlock the solutions.

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.

# Problem 1 (28 points)

An automobile consulting company wants to understand the factors on which the pricing of cars depends. The dataset `car_price_prediction` in the `math3190package` (and also found in the GitHub data folder) has information on the sales of 4340 vehicles

```{r}
#| label: loadpackages
library(math3190package)
library(glmnet)
library(glmnetUtils)
```

### Part a (3 points)

Create new variables in the dataset by taking the natural log of selling price, year, and km_driven. This will help linearize the association between variables. Then fit a linear model for predicting the log of selling price using all other variables except "name", "year", "selling_price", and "km_driven" (the logged variables should be used instread of those last three). Also, there is only one vehicle with a fuel type of "Electric". Remove that vehicle from the dataset.

```{r}
#| label: 1a
df = car_price_prediction
df = df |>
  mutate(selling_price_l=log(selling_price),
         year_l = log(year),
         km_driven_l = log(km_driven))

car_mod = lm(selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, df)

print(which(df$fuel == "Electric"))
df <- df[-4146, ]

summary(car_mod)
```

### Part b (5 points)

Now fit some LASSO models for predicting log selling price using all but the "name" variable. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ and comment on how the coefficients of the model change.

Note: when $\lambda=0$, the LASSO coefficients should equal the OLS model coefficients. However, they will actually be a bit off here. That is because the `glmnet()` function has a `thresh` argument that sets a threshold for convergence. It is, by default, set to `1e-07`. To make the parameters match, you can change that to `thresh = 1e-14` instead. This is not necessary, though.

```{r}
#| label: 1b

fit_lasso = glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 1
)

coef(fit_lasso, s=0)
coef(fit_lasso, s=0.01)
coef(fit_lasso, s=0.1)
coef(fit_lasso, s=1)
```

As lambda increases, the coefficients move closer to zero, and more variables get eliminated. All variables are eliminated at lambda = 1

### Part c (4 points)

Now fit some ridge regression models for predicting log selling price using all but name. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ and comment on how the coefficients of the model change.

```{r}
#| label: 1c
fit_ridge = glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 0
)

coef(fit_ridge, s=0)
coef(fit_ridge, s=0.01)
coef(fit_ridge, s=0.1)
coef(fit_ridge, s=1)
```

The coefficients become further from 0 as the lambda value increases. No variables are eliminated

### Part d (4 points)

Now fit some elastic net regression models for predicting log price using all but name. Try the following values for the regularization parameter: $\lambda=0,\ 0.01,\ 0.1,$ and $1$ for $\alpha=1/3$ and then comment on how the coefficients of the model change for each $\alpha$ from parts (b), (c), and (d).

```{r}
#| label: 1d
fit_elastic = glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 1/3
)

coef(fit_elastic, s=0)
coef(fit_elastic, s=0.01)
coef(fit_elastic, s=0.1)
coef(fit_elastic, s=1)
```

### Part e (4 points)

Use the `cv.glmnet()` to find the "optimal" $\lambda$ value (let's use the "lambda.1se") for LASSO ($\alpha=1$), ridge ($\alpha=0$), and elastic net (let's again use $\alpha = 1/3$) and then fit models using the `glmnet()` function for all three using their respective "best" $\lambda$. **Set a seed** before running the `cv.glmnet()` each time.

Compare each model's variables and coefficients.

```{r}
#| label: 1e

# LASSO
set.seed(7)
cvfit_lasso = cv.glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 1
)

cvfit_lasso$lambda.1se

fit_lasso = glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 1, lambda = cvfit_lasso$lambda.1se
  )

coef(fit_lasso)

# Elastic net

set.seed(7)
cvfit_elastic = cv.glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 1/3
)

cvfit_lasso$lambda.1se

fit_elastic = glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 1, lambda = cvfit_elastic$lambda.1se)

coef(fit_elastic)
  
# Ridge
set.seed(7)
cvfit_ridge = cv.glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 0
  )

cvfit_ridge$lambda.1se

fit_ridge = glmnet(
  selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df, alpha = 0, lambda = cvfit_ridge$lambda.1se
  )

coef(fit_ridge)
```

The coefficients for LASSO and elastic model is similar in value. Elastic model has fewer variables. The ridge model has all variables.

### Part f (8 points)

Use bootstrapping with at least 1000 samples to estimate the standard errors of the coefficients of the ridge regression model. For each bootstrap sample, run the `cv.glmnet()` function to find the best $\lambda$ and fit a model using that optimized $\lambda$. Use the "lambda.1se" for this. This can take a few minutes to run. Then compare these bootstrapped standard errors to the standard errors for the OLS model you fit in part a. Are they larger or smaller? Does this make sense?

```{r}
#| cache: true

bootcases = function(resids, i) {
  cvfit = cv.glmnet(selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df[i,], alpha = 0)
  fit = glmnet(selling_price_l~year_l+km_driven_l+fuel+seller_type+transmission+owner, data= df[i,], alpha=0, lambda = cvfit$lambda.1se)
  coef(fit)[,1]
}

boot_out = bootstrapping(residuals(car_mod), bootcases, 1000)
boot_out
```

Bootstrapped standard errors are smaller than the standard errors in the OLS model. This makes sense bcause ridge regression reduces variance.

# Problem 2 (21 points)

In 1990, sixty districts in California were randomly selected with the goal of predicting the median house value for the district based on the district location (longitude and latitude), the median house age in the block group, the total number of rooms in the homes, the total number of bedrooms, the population in the district, the number of households, and the median income (in \$10,000). The data are already in the .qmd file.

```{r}
#| label: house_data
#| echo: false
library(tidyverse) |> suppressPackageStartupMessages()
housing_train <- tibble(
  house_value = c(156100, 145300, 150000, 50600, 176300, 178500, 123800,
                  173900, 422700, 164400, 42100, 157800, 162900, 90500,
                  191800, 128100, 109700, 206700, 132200, 248200, 336900,
                  58000, 271500, 259500, 69400, 132700, 461200, 245000,
                  218000, 311700, 361600, 250000, 113900, 103600, 81900,
                  351900, 114100, 500001, 153300, 466700),
  longitude = c(-121.08, -121.27, -117.7, -119.29, -118.09, -117.67, 
                -120.76, -118.16, -118.39, -123.1, -117.66, -118.16, 
                -120.98, -120.99, -122.69, -118.18, -122.27, -122.11, 
                -121.99, -118.02, -118.02, -120.94, -118.6, -121.95, 
                -120.62, -118.18, -122.13, -117.93, -121.85, -118.49, 
                -122.44, -121.32, -117.07, -121.83, -121.43, -122.11, 
                -118.27, -118.4, -120.28, -117.16),
  latitude = c(38.95, 38.7, 34.06, 36.34, 33.9, 34.02, 38.47, 34.02,
               33.89, 38.79, 35.63, 33.97, 38.99, 37.67, 38.35, 33.91,
               37.83, 37.99, 38.34, 33.74, 33.73, 40.14, 34.26, 37.96, 
               36.99, 33.8, 37.4, 33.94, 36.6, 34.18, 37.76, 37.67, 32.74, 
               38, 38.53, 37.41, 34.01, 34.41, 37.9, 32.74),
  age = c(18, 16, 25, 28, 33, 16, 17, 44, 38, 20, 33, 23, 17, 28, 16, 41, 
          51, 16, 16, 26, 24, 31, 18, 18, 32, 42, 29, 30, 21, 31, 52, 21, 
          38, 15, 36, 27, 47, 22, 17, 43),
  rooms = c(1931, 3747, 2054, 1440, 3326, 3042, 1521, 1218, 1851, 3109, 
            2579, 1516, 3403, 1768, 1689, 1260, 2665, 3913, 1470, 3842, 
            6393, 3127, 6154, 2739, 2455, 2301, 6027, 2658, 2381, 3073, 
            2959, 1494, 1901, 6365, 2430, 5110, 921, 4443, 1047, 1437),
  bedrooms = c(380, 586, 609, 431, 720, 524, 309, 374, 332, 712, 564, 
               457, 661, 423, 254, 299, 574, 710, 261, 609, 1141, 664, 
               1070, 393, 508, 621, 1195, 382, 701, 674, 683, 271, 392, 
               1646, 426, 1599, 264, 560, 212, 406),
  population = c(1271, 1817, 2271, 2178, 2533, 1516, 607, 1175, 750, 1643,
                 1155, 1977, 1540, 1066, 921, 1535, 1258, 1782, 748, 1961,
                 2743, 1345, 3010, 1072, 1344, 2114, 2687, 1135, 1264, 1486,
                 1145, 781, 1099, 3838, 1199, 2764, 881, 1573, 530, 692),
  households = c(377, 590, 564, 440, 689, 475, 240, 342, 314, 638, 431, 435,
                 622, 392, 270, 322, 536, 676, 256, 595, 1057, 580, 1034, 374,
                 492, 561, 1171, 392, 659, 684, 666, 255, 406, 1458, 437,
                 1482, 221, 496, 196, 379)
)

housing_test <- tibble(
  house_value = c(214900, 44000, 194400, 128700, 435000, 179400, 129300,
                  312300, 76300, 225800, 295500, 162500, 340800, 93800,
                  155000, 283200, 133400, 248200, 265000, 57500),
  longitude = c(-119.25, -121.93, -122.31, -121.27, -118.42, -118.71, 
                -117.31, -122.04, -119.69, -117.88, -117.25, -121.48, 
                -122.49, -121.49, -116.18, -119.19, -118.23, -118.35, 
                -118.35, -124.15),
  latitude = c(34.26, 41.86, 38.27, 38.67, 33.85, 34.27, 34.02, 37.04, 
               36.79, 33.7, 32.8, 38.55, 37.74, 39.52, 33.69, 34.3, 33.89,
               34.28, 33.91, 40.88),
  age = c(30, 28, 34, 15, 43, 26, 18, 17, 15, 18, 26, 52, 48, 25, 17, 
          25, 16, 30, 31, 33),
  rooms = c(2948, 4225, 1748, 1701, 1584, 990, 1634, 4977, 2524, 2135, 
            2442, 2037, 1186, 848, 89, 2197, 5003, 3214, 2583, 2235),
  bedrooms = c(827, 835, 284, 346, 477, 223, 274, 994, 451, 373, 659, 
               358, 213, 153, 19, 320, 1180, 513, 663, 506),
  population = c(1635, 1908, 783, 723, 799, 719, 899, 1987, 1207, 1464, 
                 1134, 811, 487, 436, 79, 934, 4145, 1700, 1675, 1165),
  households = c(750, 686, 303, 352, 433, 232, 285, 947, 424, 405, 624, 
                 375, 207, 155, 21, 330, 1159, 533, 612, 441)
)
```

### Part a (2 points)

Using the training set, fit a linear regression model predicting house value from the other variables.

```{r}
#| label: 2a

lin_mod = lm(house_value ~ ., data = housing_train)
```

### Part b (4 points)

Fit a ridge regression using `cv.glmnet()` to choose the optimal $\lambda$ using "lambda.min". Set the seed before using`cv.glmnet()`.

```{r}
#| label: 2b

set.seed(7)
cvfit_ridge2 = cv.glmnet(house_value~., data=housing_train, alpha=0)
fit_ridge = glmnet(
  house_value~., data = housing_train, alpha=0, lambda= cvfit_ridge2$lambda.min
)
```

### Part c (5 points)

Compute the sum of squared errors (that is $\sum_{i=1}^n (y_i-\hat{y}_i)^2$ for both the OLS model and the ridge regression model for the **training set**. The `predict()` functions will be useful here. Compare the results and comment on why the OLS performs better here (the OLS should indeed perform better).

```{r}
#| label: 2c
# 
sse_ols_train <- sum((housing_train$house_value - predict(lin_mod))^2)
sse_ols_train

yhat_ridge_train <- predict(fit_ridge, housing_train)
sse_ridge_train <- sum((housing_train$house_value - yhat_ridge_train)^2) + 
  cvfit_ridge2$lambda.min*sum(fit_ridge$beta^2)
sse_ridge_train
```

The OLS model does perform better because we are using the training set. OLS has high variance but low bias while ridge regression has the opposite. Therefore, OLS can overfit and performs better on the training set.

### Part d (3 points)

Compute the sum of squared errors (that is $\sum_{i=1}^n (y_i-\hat{y}_i)^2$ for both the OLS model and the ridge regression model for the **testing set**. Remember to use the exact models you fit to the training sets. The `predict()` functions will be useful here. Compare the results and comment on why the ridge regression performs better here.

```{r}

#| label: 2d

sse_ols_test <- sum((housing_test$house_value - predict(lin_mod))^2)
sse_ols_test

yhat_ridge_test <- predict(fit_ridge, housing_test)
sse_ridge_test <- sum((housing_test$house_value - yhat_ridge_test)^2) + 
  cvfit_ridge2$lambda.min*sum(fit_ridge$beta^2)
sse_ridge_test
```

Ridge regression does better than OLS for the testing set because OLS can overfit while ridge regression doesn't. Ridge regression is better generalized and can stabilize when presented with noisy data.

### Part e (7 points)

Use the `train()` function in the `caret()` package to find the optimal $\alpha, \lambda$ pair in this situation (using the training set and **10-fold cross validation**). For $\lambda$, search for values between 0 and 10000 by 100 and for $\alpha$, search for values between 0 and 1 by 0.05. Then fit a model using the optimal $\alpha$ and $\lambda$ using `glmnet()` and compare the sum of squared errors using this optimized model on the test set to the two models in part (d). Set a seed before running the `train()` function.

```{r}
#| label: 2e

library(caret)
set.seed(7)

op_mod = train(house_value~., method="glmnet", 
               data = housing_train,
               tuneGrid = expand.grid(alpha = seq(0, 1, 0.05), 
                                      lambda = seq(0, 10000, 100)),
               trControl = trainControl(method = "cv", number = 10))
ggplot(op_mod, highlight=TRUE)
op_mod$bestTune

#fit model
best_mod = glmnet(house_value~., data = housing_train, alpha=0.05, lambda=3500)

# find sse
yhat_best <- predict(best_mod, housing_test)
sse_best <- sum((housing_test$house_value - yhat_best)^2) + 
  3500*(0.05*sum(best_mod$beta^2))
sse_best
```

# Problem 3 (41 points)

In the `AER` package is the data set "ShipAccidents". You can install that package and load that data using `data(ShipAccidents)`. Type `?ShipAccidents` to read about that data set.

### Part a (4 points)

Load in the data, remove the rows for which service is equal to 0, and then fit a Poisson regression model for predicting incidents from all other variables using the ShipAccidents data. Don't forget to put `family = "poisson"` in the `glm()` function. Briefly explain why it makes sense to do Poisson regression here.

```{r}
#| label: 3a

library(AER)
data(ShipAccidents)
ship = ShipAccidents

ship <- ship |>
  filter(service != 0)

pos_mod=glm(incidents~., data=ship, family="poisson")
```

The response variable is a count-based variable

### Part b (5 points)

Using that model, interpret the slope of "service" and the slope of "construction1970-74" in original (not log) units.

```{r}
#| label: 3b

summary(pos_mod)
```

As the aggregate months of service on the ship increases by 1 month, the average number of incidents is predicted to change by a factor of exp(0.06)

### Part c (6 points)

Make a residual plot of the Pearson residuals vs the linear predictors and make a QQ plot of the Pearson residuals. Comment on what these imply about the fit.

```{r}
#| label: 3c

p_resids = residuals(pos_mod, type="pearson")
resid_df = data.frame(p_resids,
                      linear_pred = pos_mod$linear.predictors)
ggplot(resid_df, aes(x=linear_pred, y=p_resids)) +
  geom_point(size=2) +
  labs(x = "Linear Predictors", y = 'Pearson Residuals') +
  theme_bw()

ggplot(resid_df, aes(sample=p_resids)) +
  geom_qq()+
  geom_qq_line(linewidth=1) +
  labs(x='Theoretical Quantiles',
       y='Sample Quantiles') +
  theme_bw()
```

The QQ plot shows that the residuals are not normally distributed, which means the model may not be a good fit. The residual plot looks pretty good.

### Part d (5 points)

Conduct a deviance goodness-of-fit (GOF) test for a lack of fit here. Type out the null and alternative hypotheses, report the test statistic, give the p-value, and provide an interpretation.

H0 = The model is correctly specified

HA = The model is incorrectly specified

Residual deviance = 77.45

p-value = 0.00000015

We have strong evidence that the model is incorrectly specified.

```{r}
#| label: 3d

1 - pchisq(pos_mod$deviance, df = pos_mod$df.residual)
```

### Part e (4 points)

Is there evidence that this model has overdispersion? Explain.

Yes, one of the predictors are not significant.

### Part f (6 points)

Fit a "quasipoisson" model to these data. Using this model, obtain an approximate 95% confidence interval for the mean response for a ship of type "B" with construction between 1965 and 1969, with operation between 1960 and 1974, and with 2000 aggregate months of service. You can construct this interval using a t critical value with 24 degrees of freedom since the quasipoisson family more closely follows a t distribution than a normal one. Interpret the interval.

```{r}
#| label: 3f

quasi_mod = glm(formula = incidents ~., family = "quasipoisson", data=ship)

summary(quasi_mod)

ship_pred = predict(quasi_mod, 
                    data.frame(type="B",
                               construction = "1965-69",
                               operation = "1960-74",
                               service = 2000),
                    se.fit=T)

tcrit = qt(0.975, df = 24)
tcrit

exp(ship_pred$fit - tcrit * ship_pred$se.fit)
exp(ship_pred$fit + tcrit * ship_pred$se.fit)
```

We are 95% confident that the true average number of incidents for a ship of type "B" with construction between 1965 and 1969, with operation betwen 1960 and 1974, and with 2000 aggregate months of service is between 3.66 and 19.95.

### Part g (5 points)

Now construct an approximate 95% prediction interval for the number of incidents for the ship described in part f using the `qpois()` function. Interpret the interval.

```{r}
#| label: 3g

qpois(0.05, 3.66)
qpois(0.95, 19.95)
```

We are 95% confident that the true number of incidents for a ship of type "B" with construction between 1965 and 1969, with operation betwen 1960 and 1974, and with 2000 aggregate months of service is between 1 and 28.

### Part h (6 points)

Suppose we did not catch the fact that this model is overdispersed. Repeat parts f and g using the model fit with the "poisson" family, not "quasipoisson" and using a z interval for the mean response rather than a t interval. Compare your results to parts f and g and comment on what is different.

```{r}
#| label: 3h

ship_pred = predict(pos_mod, 
                    data.frame(type="B",
                               construction = "1965-69",
                               operation = "1960-74",
                               service = 2000),
                    se.fit=T)


exp(ship_pred$fit - 1.96 * ship_pred$se.fit)
exp(ship_pred$fit + 1.96 * ship_pred$se.fit)
```

We are 95% confident that the true average number of incidents for a ship of type "B" with construction between 1965 and 1969, with operation betwen 1960 and 1974, and with 2000 aggregate months of service is between 5.33 and 13.71.

```{r}

#| label: 3h2

qpois(0.05, 5.33)
qpois(0.95, 13.71)
```

We are 95% confident that the true number of incidents for a ship of type "B" with construction between 1965 and 1969, with operation betwen 1960 and 1974, and with 2000 aggregate months of service is between 2 and 20.

The confidence/prediction interval for the normal Poisson model is smaller than that of the quasi-Poisson model.
